Interpretable Neural Networks using EAGGA

### 2025-03-15 paper/main.tex
experimental results + discussion

The dominated hypervolume over generations in Figure \ref{fig-val-dhvs}

- dhv over generations Figure \ref{fig-val-dhvs}
  * compute on val set, i.e. had 5 folds per individual
    - NF, NI, NNM always the same for each fold
    - but 5 different AUCs
    -> hypervolume of (mean(AUC 1, AUC 2, ..., AUC 5), NF, NI, NNM)
  * artifacts / drops along y likely due to inconsistent computation of Pareto front by third-party library
    - preliminary experiments on dummy pareto fronts with 10x4 metrics: noticed nds function returning different rankings for same front
    - returned ranking switched back and forth between two only slightly different options (only 1 or 2 indices were swapped)
    - not sure what caused this, as made same observation using another library that was planned as alternative
    - thus unfortunately not fixable for me
  * all but 3 datasets (2 of which only trained for 1 generation, aynway) show improvement of dominated hypervolume over generations
  * BUT: absolute as well as relative improvement almost negligible -> no large final dhv decrease would we just have evaluated the models gotten from the detectors
    - this was also the reason we didn't run EAGGA on philippine and gina simply without the detectors (as that's where they crashed) -> evidence suggested that detectors are vital to initial performance
    - original paper supports this assumption \citep[Fig. 4, p. 545]{EAGGA}

    - preliminary
  * our implementation of feature + interaction detectors likely tend to include  % belongs in "general algorithm"
    - more features + interactions for high p datasets
    - less for low p datasets
    - remember: original paper samples \# of features included from truncated geometric distribution, what is not mentioned is probability of this distribution
      is determined from fitting 10 trees + looking at relative \# of features used (cf. their github repo ~/R/TunerEAGGA.R function get\_n\_selected\_rpart())
    -> mlr3 default decision tree max depth 30
      * i.e. datasets with p < 30 might use all features, which translates to trunc geom prob = 1
      * vice-versa for p >> 30 datasets relative \# features < 0.5 (our trunc geom prob)
      * similar reasoning for pairwise interactions
  * individuals not in Pareto sets occassionally exhibit AUC < 0.5
    - this is not by mistake, usually AUC on binary target can simply be inverted (simply predict the opposite class)
    - for us this cannot easily be inverted due to monotonicity constraints (weights cannot simply be multiplied by -1 if constrained)
    - could happen e.g. due to AdamW and weight clipping (employed to enforce monotonicity)
      * AdamW uses momentum
      * possible scenario: large momentum would yield negative weights, but after epoch they're clipped to [0,infty) + training stops due to early stopping
      * weight clipping very imperfect way of enforcing monotonicity, but currently in pytorch unfortunately only way to implement this

@@

\section{Plots}
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{../code/export/plot_early_stopping_losses_pareto_set.png}
  \caption{Pareto set loss of all datasets, evaluated on early stopping set. Same colours denote losses coming from folds of the same individual.
            x-axis portrait epochs, y-axis binary cross entropy loss.}
  \label{fig-es-losses}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{../code/export/plot_val_dhvs_over_time.png}
  \caption{Dominated hypervolume over generations, evaluated on validation set.
            x-axis portrait generations, y-axis dominated hypervolume using mean AUC over folds.}
  \label{fig-val-dhvs}
\end{figure}

>>>>

The AUC used in computing the dominated hypervolume over generations in Figure \ref{fig-val-dhvs} is the mean AUC over all folds on the validation set.
All but 3 datasets, 2 of which were only trained for 1 generation, exhibit improvement of their dominated hypervolume over generations. The absolute as
well as relative improvement, however, is almost negligible, suggesting there to be no large decrease in dominated hypervolume, would we have just evaluated
the group structure initialised by the detectors.
This was also the reason behind not running EAGGA on the philipine and gina datasets on randomly initialised detectors (as the detectors for those datasets
ran out of memory due to the high feature cardinality), as the detectors prove vital for the initial performance. This finding is also in
line with \citep[Fig. 4, p. 545]{EAGGA}.
Clearly, there are some artifacts causing drops along the y-axis in most plots. This is likely due to the inconsistent computation of the Pareto front
by a third-party library: in preliminary experiments we observed slightly varying front ranks on a synthetic sample of 10 (AUC, NF, NI, NNM) vectors.
Unfortunately, this behaviour was still present when switching to another library for front rank computation, and thus not fixable for us.
From our anecdotal evidence, the variations in rankings in these preliminary experiments only concerned few samples that were assigned to alternating
neighbouring fronts, which seemed reasonable not to be concerned over.  % ok formulation?
Another observation from the logs during training is AUC occassionally being less than 0.5.
In general, AUC can then simply be inverted, as the model can usually always simply predict the opposite.
In our case, however, this is not possible as easily due to weight clipping for groups with a monotonicty constraint.
A potential scenario were an AUC less than 0.5 could happen is when AdamW has high momentum, leading to negative weights that are then clipped to the
non-negative domain at the end of the epoch.
We acknowledge that weight clipping is a very imperfect way of enforcing monotonicity, but currently seems to be the only way to implement this type of
constrained optimisation in pytorch.

% belongs in "general algorithm"
%    - preliminary
%  * our implementation of feature + interaction detectors likely tend to include
%    - more features + interactions for high p datasets
%    - less for low p datasets
%    - remember: original paper samples \# of features included from truncated geometric distribution, what is not mentioned is probability of this distribution
%      is determined from fitting 10 trees + looking at relative \# of features used (cf. their github repo ~/R/TunerEAGGA.R function get\_n\_selected\_rpart())
%    -> mlr3 default decision tree max depth 30
%      * i.e. datasets with p < 30 might use all features, which translates to trunc geom prob = 1
%      * vice-versa for p >> 30 datasets relative \# features < 0.5 (our trunc geom prob)
%      * similar reasoning for pairwise interactions

@@

\section{Plots}
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{../code/export/plot_early_stopping_losses_pareto_set.png}
  \caption{Pareto set loss of all datasets, evaluated on early stopping set. Same colours denote losses coming from folds of the same individual.
            x-axes portrait epochs, y-axes binary cross entropy loss.}
  \label{fig-es-losses}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{../code/export/plot_val_dhvs_over_time.png}
  \caption{Dominated hypervolume over generations, evaluated on validation set.
            x-axes portrait generations, y-axes dominated hypervolume using mean AUC over folds.}
  \label{fig-val-dhvs}
\end{figure}
